Why do people punish? Evidence for a range of strategic concerns

### 2024-05-22 manuscript.Rmd
submit manuscript

keywords          : "punishment; cooperation; economic games"
wordcount         : "5657 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

@@

\nolinenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost[@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target (or bystanders 
[@DosSantos2011;@DosSantos2013;@Raihani2015]) in the future. Punishment offers a 
route to maintaining or increasing cooperation by changing the payoff structure 
of social interactions such that it no longer pays to cheat or exploit social 
partners[@CluttonBrock1995;@Ostrom1990]. Yet, despite its theoretical 
importance, the question of why people choose to punish others is still hotly 
contested[@Raihani2019]. In this study, we use a battery of economic games to 
disentangle the different motives underpinning punishment and explore how these 
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory 
settings using economic games[@Balliet2011;@Balliet2013;@Chaudhuri2011;
@Dreber2008;@Fehr2000;@Fehr2002;@Henrich2006;@Nikiforakis2008a;@Raihani2012a;
@Raihani2019]. In these games, participants are usually given a sum of money 
that they can use to invest in collective action or to help others. 
Alternatively, participants can 'cheat' by keeping the money for themselves or 
by exploiting the contributions of others. Punishment is introduced into such 
games by giving participants the option to pay a small 'fee' to impose a greater 
'fine' on their co-players. Several lines of experimental evidence indicate that 
people use this punishment option[@Henrich2006], that they enjoy 
punishing[@DeQuervain2004], and that they frequently, though not 
always[@Herrmann2008], punish cheating or exploitative 
co-players[@Fehr2000;@Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment 
plays an important role in promoting human cooperation. People tend to cooperate 
more in games where punishment is possible compared to those where it is not
[@Balliet2011;@Balliet2013;@Raihani2019]. The effect that the threat of 
punishment has on cooperation is also evident in the higher contributions 
typically observed in the Ultimatum Game (where punishment is possible) compared 
to the structurally-similar Dictator Game (where it is not)[@Camerer2003]. This 
typical cooperation-enhancing effect of punishment has also been observed across 
societies[@Balliet2013], leading some to suggest that costly punishment has 
played a key role in the cultural evolution of cooperation in humans
[@Bowles2013;@Boyd2003;@Chudek2011;@Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to 
achieve other ends. Some have argued that punishment is primarily used to shape 
future behaviour, either to deter personal harm[@Delton2017;@Dreber2008;
@DosSantos2013] or to uphold normative standards of cooperative behaviour 
[@Boyd2003;@Chudek2011;@Fehr2018;@Mathew2011;@Mathew2014;@Richerson2016]. 
But while the _threat_ of punishment can have a cooperation-enhancing effect, 
the _enactment_ of this punishment does not consistently deter targets from 
cheating in the future[@Raihani2019]. This calls into question whether 
punishment primarily operates as a behaviour-change tool or whether it is used 
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused[@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others[@Raihani2012b;@Raihani2019]) and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs[@Dawes2007]). Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position[@Raihani2019].

Common economic game designs have been unable to tease apart these different 
motives for punishment because participants who interact with cheats in these 
games experience both losses _and_ lower relative payoffs. The typical 1:3 
fee-fine ratio of punishment in economic games compounds this issue. With 
this setup, people can simultaneously use punishment to reciprocate losses, to 
deter others from cheating, and to reduce or reverse disparities in payoffs 
between themselves and targets. To add to this complexity, it is evident that 
people use punishment in seemingly disparate ways: punishing when no behaviour 
change is possible, such as in one-shot games[@Bone2015;@Henrich2006;
@Raihani2012b;@Walker2004], on the very last round of repeated games
[@Barclay2016], or in games where the target never learns about the 
punishment[@Crockett2014]; punishing those who did not cheat or who 
over-contributed to collective action (antisocial punishment
[@Herrmann2008;@Sylwester2013]); punishing in scenarios where they were not 
personally harmed (third-party punishment[@Fehr2004]); and punishing in 
scenarios where disparities in payoffs did not arise from participants' 
actions[@Dawes2007;@Johnson2009;@Fowler2005].

The general conclusion from this research is that there is no one unifying 
function of costly punishment in humans. Instead, punishment should be thought 
of as a flexible behavioural tool that serves a variety of functions that are 
not mutually exclusive[@Raihani2019]. Due to its multipurpose nature, we should 
therefore expect variation in punishment strategies in the population, much like 
the observed variation in social learning strategies[@Molleman2014]. Some 
individuals may use punishment as a behaviour shaping tool, for example, while 
others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment 
strategies are more frequent in human populations? Second, what traits predict 
adherence to a particular strategy? Previous work has reported that personality 
is related to cooperative behaviour[@Thielmann2020] and demographics, political 
ideology, and religiosity are related to punitive behaviour[@Hofmann2018], but 
no research has related these variables to specific punishment strategies. 
Third, do people have insight into their own punishment strategy? Previous work 
has argued that people are often unaware of the underlying function of their 
punitive behaviour, yet they feel compelled to enact it anyway
[@Carlsmith2002;@Nisbett1977].

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness[@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

Building on previous designs[@Bone2015;@Deutchman2021;@Marczyk2017;
@Raihani2012b], we employ a suite of one-shot economic games where individuals 
are given the opportunity to punish targets at a personal cost (Figure 
\@ref(fig:plotGames)). In each game, targets either steal from another 
individual or do nothing. Representative samples of participants from the United 
Kingdom (*n* = `r sum(tar_read(dExc)$Country == "United Kingdom")`) and the 
United States (*n* = `r sum(tar_read(dExc)$Country == "United States")`) 
completed all six games on the online platform Prolific. We carefully designed 
the suite of games to tease apart the proposed punishment strategies in Table 
\@ref(tab:tableStrategies), such that each strategy predicts a different pattern 
of behaviour across all the games (see Methods for more detail about the six 
games). We use the resulting behavioural patterns to discern which punishment 
strategy participants are employing. We then combine these behavioural patterns 
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

@@

Using a suite of economic games measuring punishment in different situations, 
we have shown that punishment does not serve just one function, but instead is 
a flexible tool that can be and is used for different purposes[@Raihani2019]. 
While some use punishment to enforce norms of cooperation, others use it to 
reduce or even create inequality between individuals. We found that people's 
punishment strategy can, to some extent, be predicted by individual differences 
in personality, social preferences, and political and religious views. Moreover, 
contrary to the view that people are often unable to articulate the reasons for 
their punitive behaviour[@Carlsmith2002], people seem to have some degree of 
insight into the strategy they are using. Despite small differences, these 
general patterns replicated in samples from both the United Kingdom and the 
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to 
inequality in payoffs, either from a self-referential perspective (i.e., avoid 
disadvantageous inequity) or more generally (i.e., egalitarian). This is in 
line with previous studies which have highlighted inequity aversion as an 
important motivation for punishment in economic games 
[@Bone2015;@Bone2016;@Dawes2007;@Deutchman2021;@Johnson2009;@Raihani2012b]. 
Personality, social preferences, and political and religious views were related 
to these strategies. Traits associated with other-regarding concern, such as SVO 
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to 
follow the avoid disadvantageous inequity strategy, especially in the United 
States. Authoritarian participants in the United States were also more likely to 
actively seek advantageous inequity with their punishment. These findings 
provide insight into the motives that may have driven higher levels of peer 
punishment among conservatives in previous work[@Hofmann2018].

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less 
common than strategies sensitive to inequality in our set of games. This was 
reflected both in participants’ elicited punishment behaviour (Figure 
\@ref(fig:plotModel1b)) and in their self-reports of their own strategy 
(Supplementary Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2)). 
Although our design did not explicitly allow for behaviour shaping as the 
interactions were all one-shot, we did manipulate whether the target's stealing 
behaviour was intentional or not (Game C), an approach which has been used in 
previous vignette studies to identify behaviour shaping motives[@Carlsmith2002].
The lower prevalence of behaviour shaping strategies in our study is consistent 
with prior work showing that punishment often continues to be used when 
behaviour shaping is impossible, such as when the target will never find out 
that they have been punished[@Crockett2014] or on the last round of repeated 
interactions[@Barclay2016]. We found that demographic, personality, political, 
and religious variables tended to be unrelated to behaviour shaping strategies, 
though this could reflect low power to detect such associations given the low 
prevalence of these strategies in our sample. We also found that participants 
accurately reported using the norm-enforcing strategy, but not the deterrent 
strategy. This finding is in line with previous research showing that people 
struggle to accurately report the deterrent motivations for their punitive 
behaviour[@Carlsmith2002].

We defined antisocial punishment as occurring when a participant exclusively 
punished those who did not steal. There were no participants in our sample who 
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the 
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all 
games suggests that the antisocial punishment that has been observed in other 
studies[@Herrmann2008;@Pleasant2018] is not aimed at harming cooperators 
specifically, as has been previously suggested[@Herrmann2008]. Instead, 
antisocial punishment is more likely to be motivated by improving one's 
relative position, which is in line with work showing that antisocial punishment 
disappears when relative payoffs cannot be changed (e.g., when punishment is 
only available with a 1:1 fee-fine ratio[@Sylwester2013]).

The fact that people use punishment for many different reasons poses problems 
for the way that punishment is operationalised in classic behavioural economic 
game studies. In these studies, a common assumption is that participants punish 
to change the behaviour of cheats[@Fehr2000;@Fehr2002]. But in reality, people 
may be choosing the punishment option to achieve a variety of different goals. 
This has implications for how people respond to being targeted by punishers in 
these games. Targets of punishment in these studies may know that punishment 
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation[@Dreber2008;@Nikiforakis2008b;@Raihani2019].
As punishers’ motives must be inferred (and such inferences likely depend on 
character traits of the target, as well as the context in which punishment 
occurs), there is likely to be some variation and error in attributing motives 
to punishers. To the extent that inferred motives affect target responses, this 
might help to explain the mixed findings in the field as to whether punishment 
actually motivates cheating targets to cooperate in the future[@Raihani2019].

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies[@Baumard2010]. 
One reason that people may be averse to peer punishment is that, because it is 
a fundamentally harmful act, punishment can reflect badly on the punisher and 
people might therefore refrain from punishing others to avoid reputational 
damage[@Raihani2015]. People frequently avoid taking actions that could harm 
their reputation, even when they don’t know if reputation is at stake (as in the 
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation[@Raihani2019]. This may be
especially likely in situations that lack clear institutional norms to 
legitimise punishment, such as our economic games. As with reputation damage, 
people might abstain from peer punishment to avoid retaliation, regardless of 
whether retaliation is actually possible. By contrast, institutionalised 
punishment in small-scale societies often functions to compensate victims while 
limiting the potential for feuds and cycles of retaliation
[@Fitouchi2023;@Singh2022]. Future research should uncover whether people are 
more willing to punish in these conventionalised contexts (e.g., see
[@Molleman2019]).

@@

Another limitation is that our results may be contingent on the particular suite 
of anonymous stealing games that we used. With our anonymous design, we were 
unable to study other potential reputational strategies underlying punishment, 
such as signalling trustworthiness[@Jordan2020]. Moreover, stealing may be 
evaluated differently to other forms of cheating, such as not contributing to 
public goods, and other negative behaviours, such as lying or breaking taboos. 
Finally, we were unable to include all permutations of situational features in 
the games (e.g., second-party vs. third-party, equal vs. unequal) making it 
difficult to interpret some patterns of behaviour. For example, many 
participants punished only in the third-party game (Game F), but it is not clear 
whether these participants were driven by the third-party nature of the game or 
simply by the fact that stealing in that game generated inequality. Future work 
should determine whether different reputational contexts, target behaviours, and 
combinations of situational features elicit different punishment strategies.

In sum, we have shown that while many people choose not to punish peers, those 
who do are motivated by a variety of different concerns, including behaviour 
shaping, egalitarianism, and competition. Much like the observed variation in 
human social learning strategies[@Molleman2014], humans thus also exhibit 
variation in their punishment strategies. These individual differences map onto 
personality dimensions, social preferences, political and religious views, and 
self-reports of behaviour. We hope that future work will continue to unpack the 
multifaceted nature of human punishment.

@@

We pre-registered that we would use a Bayesian latent state model to infer 
unobserved punishment strategies from the observed data (for a similar version 
of this model, see ref[@McElreath2020]). In this model, participants $i$ in 
countries $c$ make binary punishment decisions across twelve decisions $j$. We 
assume that the probability of the observed data $y_{i,j}$ is the weighted 
average of the probability of the observed data conditional on each of the ten 
punishment strategies $s$. From this logic, the model estimates the probability 
of each strategy $p_{s}$. The full model is as follows:

>>>>

bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

@@

# Significance Statement

Research has shown that humans are willing to punish others at a personal cost,
but the motives for this behaviour remain unclear. Using a suite of economic 
games, we show that punitive behaviour is consistent with a variety of strategic
concerns. While most people choose not to punish in our games, people who do 
punish tend to be motivated by egalitarianism rather than deterrence or 
norm enforcement. We also find that different punishment strategies are linked 
with different character traits, including personality, prosocial preferences, 
and political and religious views. This research highlights the multifaceted
nature of human punishment.

\newpage
\linenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost [@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target, or bystanders 
[@DosSantos2011;@DosSantos2013;@Raihani2015], in the future. Punishment offers a 
route to maintaining or increasing cooperation by changing the payoff structure 
of social interactions such that it no longer pays to cheat or exploit social 
partners [@CluttonBrock1995;@Ostrom1990]. Yet, despite its theoretical 
importance, the question of why people choose to punish others is still hotly 
contested [@Raihani2019]. In this study, we use a battery of economic games to 
disentangle the different motives underpinning punishment and explore how these 
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory 
settings using economic games [@Balliet2011;@Balliet2013;@Chaudhuri2011;
@Dreber2008;@Fehr2000;@Fehr2002;@Henrich2006;@Nikiforakis2008a;@Raihani2012a;
@Raihani2019]. In these games, participants are usually given a sum of money 
that they can use to invest in collective action or to help others. 
Alternatively, participants can 'cheat' by keeping the money for themselves or 
by exploiting the contributions of others. Punishment is introduced into such 
games by giving participants the option to pay a small 'fee' to impose a greater 
'fine' on their co-players. Several lines of experimental evidence indicate that 
people use this punishment option [@Henrich2006], that they enjoy 
punishing [@DeQuervain2004], and that they frequently, though not 
always [@Herrmann2008], punish cheating or exploitative 
co-players [@Fehr2000;@Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment 
plays an important role in promoting human cooperation. People tend to cooperate 
more in games where punishment is possible compared to those where it is not
[@Balliet2011;@Balliet2013;@Raihani2019]. The effect that the threat of 
punishment has on cooperation is also evident in the higher contributions 
typically observed in the Ultimatum Game (where punishment is possible) compared 
to the structurally-similar Dictator Game (where it is not) [@Camerer2003]. This 
typical cooperation-enhancing effect of punishment has also been observed across 
societies [@Balliet2013], leading some to suggest that costly punishment has 
played a key role in the cultural evolution of cooperation in humans
[@Bowles2013;@Boyd2003;@Chudek2011;@Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to 
achieve other ends. Some have argued that punishment is primarily used to shape 
future behaviour, either to deter personal harm [@Delton2017;@Dreber2008;
@DosSantos2013] or to uphold normative standards of cooperative behaviour 
[@Boyd2003;@Chudek2011;@Fehr2018;@Mathew2011;@Mathew2014;@Richerson2016]. 
But while the _threat_ of punishment can have a cooperation-enhancing effect, 
the _enactment_ of this punishment does not consistently deter targets from 
cheating in the future [@Raihani2019]. This calls into question whether 
punishment primarily operates as a behaviour-change tool or whether it is used 
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused [@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others) [@Raihani2012b;@Raihani2019] and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs) [@Dawes2007]. Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position [@Raihani2019].

Common economic game designs have been unable to tease apart these different 
motives for punishment because participants who interact with cheats in these 
games experience both losses _and_ lower relative payoffs. The typical 1:3 
fee-fine ratio of punishment in economic games compounds this issue. With 
this setup, people can simultaneously use punishment to reciprocate losses, to 
deter others from cheating, and to reduce or reverse disparities in payoffs 
between themselves and targets. To add to this complexity, it is evident that 
people use punishment in seemingly disparate ways: punishing when no behaviour 
change is possible, such as in one-shot games [@Bone2015;@Henrich2006;
@Raihani2012b;@Walker2004], on the very last round of repeated games
[@Barclay2016], or in games where the target never learns about the 
punishment [@Crockett2014]; punishing those who did not cheat or who 
over-contributed to collective action (antisocial punishment) 
[@Herrmann2008;@Sylwester2013]; punishing in scenarios where they were not 
personally harmed (third-party punishment) [@Fehr2004]; and punishing in 
scenarios where disparities in payoffs did not arise from participants' 
actions [@Dawes2007;@Johnson2009;@Fowler2005].

The general conclusion from this research is that there is no one unifying 
function of costly punishment in humans. Instead, punishment should be thought 
of as a flexible behavioural tool that serves a variety of functions that are 
not mutually exclusive [@Raihani2019]. Due to its multipurpose nature, we should 
therefore expect variation in punishment strategies in the population, much like 
the observed variation in social learning strategies [@Molleman2014]. Some 
individuals may use punishment as a behaviour shaping tool, for example, while 
others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment 
strategies are more frequent in human populations? Second, what traits predict 
adherence to a particular strategy? Previous work has reported that personality 
is related to cooperative behaviour [@Thielmann2020] and demographics, political 
ideology, and religiosity are related to punitive behaviour [@Hofmann2018], but 
no research has related these variables to specific punishment strategies. 
Third, do people have insight into their own punishment strategy? Previous work 
has argued that people are often unaware of the underlying function of their 
punitive behaviour, yet they feel compelled to enact it anyway
[@Carlsmith2002;@Nisbett1977].

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness [@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives [but see [@Jordan2020]].

Building on previous designs [@Bone2015;@Deutchman2021;@Marczyk2017;
@Raihani2012b], we employ a suite of one-shot economic games where individuals 
are given the opportunity to punish targets at a personal cost (Figure 
\@ref(fig:plotGames)). In each game, targets either steal from another 
individual or do nothing. Representative samples of participants from the United 
Kingdom (*n* = `r sum(tar_read(dExc)$Country == "United Kingdom")`) and the 
United States (*n* = `r sum(tar_read(dExc)$Country == "United States")`) 
completed all six games on the online platform Prolific. We carefully designed 
the suite of games to tease apart the proposed punishment strategies in Table 
\@ref(tab:tableStrategies), such that each strategy predicts a different pattern 
of behaviour across all the games (see Methods for more detail about the six 
games). We use the resulting behavioural patterns to discern which punishment 
strategy participants are employing. We then combine these behavioural patterns 
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

@@

Using a suite of economic games measuring punishment in different situations, 
we have shown that punishment does not serve just one function, but instead is 
a flexible tool that can be and is used for different purposes [@Raihani2019]. 
While some use punishment to enforce norms of cooperation, others use it to 
reduce or even create inequality between individuals. We found that people's 
punishment strategy can, to some extent, be predicted by individual differences 
in personality, social preferences, and political and religious views. Moreover, 
contrary to the view that people are often unable to articulate the reasons for 
their punitive behaviour [@Carlsmith2002], people seem to have some degree of 
insight into the strategy they are using. Despite small differences, these 
general patterns replicated in samples from both the United Kingdom and the 
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to 
inequality in payoffs, either from a self-referential perspective (i.e., avoid 
disadvantageous inequity) or more generally (i.e., egalitarian). This is in 
line with previous studies which have highlighted inequity aversion as an 
important motivation for punishment in economic games 
[@Bone2015;@Bone2016;@Dawes2007;@Deutchman2021;@Johnson2009;@Raihani2012b]. 
Personality, social preferences, and political and religious views were related 
to these strategies. Traits associated with other-regarding concern, such as SVO 
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to 
follow the avoid disadvantageous inequity strategy, especially in the United 
States. Authoritarian participants in the United States were also more likely to 
actively seek advantageous inequity with their punishment. These findings 
provide insight into the motives that may have driven higher levels of peer 
punishment among conservatives in previous work [@Hofmann2018].

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less 
common than strategies sensitive to inequality in our set of games. This was 
reflected both in participants’ elicited punishment behaviour (Figure 
\@ref(fig:plotModel1b)) and in their self-reports of their own strategy 
(Supplementary Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2)). 
Although our design did not explicitly allow for behaviour shaping as the 
interactions were all one-shot, we did manipulate whether the target's stealing 
behaviour was intentional or not (Game C), an approach which has been used in 
previous vignette studies to identify behaviour shaping motives [@Carlsmith2002].
The lower prevalence of behaviour shaping strategies in our study is consistent 
with prior work showing that punishment often continues to be used when 
behaviour shaping is impossible, such as when the target will never find out 
that they have been punished [@Crockett2014] or on the last round of repeated 
interactions [@Barclay2016]. We found that demographic, personality, political, 
and religious variables tended to be unrelated to behaviour shaping strategies, 
though this could reflect low power to detect such associations given the low 
prevalence of these strategies in our sample. We also found that participants 
accurately reported using the norm-enforcing strategy, but not the deterrent 
strategy. This finding is in line with previous research showing that people 
struggle to accurately report the deterrent motivations for their punitive 
behaviour [@Carlsmith2002].

We defined antisocial punishment as occurring when a participant exclusively 
punished those who did not steal. There were no participants in our sample who 
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the 
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all 
games suggests that the antisocial punishment that has been observed in other 
studies [@Herrmann2008;@Pleasant2018] is not aimed at harming cooperators 
specifically, as has been previously suggested [@Herrmann2008]. Instead, 
antisocial punishment is more likely to be motivated by improving one's 
relative position, which is in line with work showing that antisocial punishment 
disappears when relative payoffs cannot be changed (e.g., when punishment is 
only available with a 1:1 fee-fine ratio) [@Sylwester2013].

The fact that people use punishment for many different reasons poses problems 
for the way that punishment is operationalised in classic behavioural economic 
game studies. In these studies, a common assumption is that participants punish 
to change the behaviour of cheats [@Fehr2000;@Fehr2002]. But in reality, people 
may be choosing the punishment option to achieve a variety of different goals. 
This has implications for how people respond to being targeted by punishers in 
these games. Targets of punishment in these studies may know that punishment 
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation [@Dreber2008;@Nikiforakis2008b;@Raihani2019].
As punishers’ motives must be inferred (and such inferences likely depend on 
character traits of the target, as well as the context in which punishment 
occurs), there is likely to be some variation and error in attributing motives 
to punishers. To the extent that inferred motives affect target responses, this 
might help to explain the mixed findings in the field as to whether punishment 
actually motivates cheating targets to cooperate in the future [@Raihani2019].

It is striking that the most common strategy in our dataset was to never punish. 
This is partly because punishment in these games imposes an economic cost for no 
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper 
to punish, we may have seen more punishment from participants. Indeed, 
`r printnum(mean(tar_read(dExc)$SelfRate10[tar_read(dExc)$strategy == "Never punish"] > 0, na.rm = TRUE)*100, digits = 0)`% 
of participants following the never punish strategy positively stated that they 
didn't want to pay to reduce anyone's bonus but would have done so if it were 
free. But the frequency of the never punish strategy perhaps also reflects a 
more general aversion to peer punishment, an aversion that has been highlighted 
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples
[@Balafoutas2014;@Balafoutas2016] and in small-scale societies [@Baumard2010]. 
One reason that people may be averse to peer punishment is that, because it is 
a fundamentally harmful act, punishment can reflect badly on the punisher and 
people might therefore refrain from punishing others to avoid reputational 
damage [@Raihani2015]. People frequently avoid taking actions that could harm 
their reputation, even when they don’t know if reputation is at stake (as in the 
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation [@Raihani2019]. This may be
especially likely in situations that lack clear institutional norms to 
legitimise punishment, such as our economic games. As with reputation damage, 
people might abstain from peer punishment to avoid retaliation, regardless of 
whether retaliation is actually possible. By contrast, institutionalised 
punishment in small-scale societies often functions to compensate victims while 
limiting the potential for feuds and cycles of retaliation
[@Fitouchi2023;@Singh2022]. Future research should uncover whether people are 
more willing to punish in these conventionalised contexts [e.g., see
[@Molleman2019]].

@@

Another limitation is that our results may be contingent on the particular suite 
of anonymous stealing games that we used. With our anonymous design, we were 
unable to study other potential reputational strategies underlying punishment, 
such as signalling trustworthiness [@Jordan2020]. Moreover, stealing may be 
evaluated differently to other forms of cheating, such as not contributing to 
public goods, and other negative behaviours, such as lying or breaking taboos. 
Finally, we were unable to include all permutations of situational features in 
the games (e.g., second-party vs. third-party, equal vs. unequal) making it 
difficult to interpret some patterns of behaviour. For example, many 
participants punished only in the third-party game (Game F), but it is not clear 
whether these participants were driven by the third-party nature of the game or 
simply by the fact that stealing in that game generated inequality. Future work 
should determine whether different reputational contexts, target behaviours, and 
combinations of situational features elicit different punishment strategies.

In sum, we have shown that while many people choose not to punish peers, those 
who do are motivated by a variety of different concerns, including behaviour 
shaping, egalitarianism, and competition. Much like the observed variation in 
human social learning strategies [@Molleman2014], humans thus also exhibit 
variation in their punishment strategies. These individual differences map onto 
personality dimensions, social preferences, political and religious views, and 
self-reports of behaviour. We hope that future work will continue to unpack the 
multifaceted nature of human punishment.

@@

We pre-registered that we would use a Bayesian latent state model to infer 
unobserved punishment strategies from the observed data [for a similar version 
of this model, see ref [@McElreath2020]]. In this model, participants $i$ in 
countries $c$ make binary punishment decisions across twelve decisions $j$. We 
assume that the probability of the observed data $y_{i,j}$ is the weighted 
average of the probability of the observed data conditional on each of the ten 
punishment strategies $s$. From this logic, the model estimates the probability 
of each strategy $p_{s}$. The full model is as follows:
